Lattices of acyclic pipe dreams

### 2023-09-25 acyclicFacets/draft/latticePipeDreams.tex
new proposition proof

This paper develops a similar framework for acyclic pipe dreams.
Pipe dreams were introduced by N.~Bergeron and S.~Billey in~\cite{BergeronBilley} to compute Schubert polynomials and later revisited in the context of Gr\"obner geometry by A.~Knutson and E.~Miller~\cite{KnutsonMiller-GroebnerGeometry}, who coined the name \emph{pipe dreams}.
They have important connections and applications to various areas related to Schubert calculus and Schubert varieties~\cite{LascouxSchutzenberger-PolynomesSchubert, LascouxSchutzenberger-SchubertLittlewoudRichardson}. 
A pipe dream is an arrangement of pipes in the triangular shape, each entering along the vertical side and exiting along the horizontal side (see \cref{fig:pipeDreams}).
They are grouped according to their exiting permutation, given by the order in which the pipes appear along the horizontal axis.
The linear extensions of a pipe dream are the permutations of its pipes such that for each contact the northwest pipe appears before the southeast pipe in the permutation.
The pipe dreams with at least one linear extension are called acyclic and naturally appear in the study of brick polytopes~\cite{PilaudSantos-brickPolytope}.
A flip in a pipe dream exchanges a contact with a crossing between two pipes (see \cref{fig:pipeDreams}), and the flip is increasing when the contact is southwest of the crossing involved in the flip.
A brief recollection on pipe dreams is given in \cref{sec:preliminaries}.

@@

The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$ in matrix notation.
A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrers diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding.
For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~${0, 1, \dots, n}$, and we consider here pipe dreams with pipes indexed by~${0, 1, \dots, n}$.
The $\nu$-Tamari lattice is the increasing flip poset on pipe dreams~$\pipeDreams(0\omega_\nu)$ for some dominant permutation $\omega_\nu$ determined by $\nu$, see~\cite{CeballosPadrolSarmiento}.

The main example is when $\omega=n\dots21$ is the reverse permutation, see \cref{exm:Tamari1}.
In this case, the increasing flip poset on~$\pipeDreams(0 \omega)$ is the classical Tamari lattice. Removing vertex $0$ from the contact graph of a pipe dream~$ P\in \pipeDreams(0 \omega)$ returns the binary tree corresponding to $P$ (with the leaves removed and the internal nodes labeled in inorder from $1$ to $n$).
The edges of the binary tree are oriented going away from the root of the tree.
See \cref{fig:bijection}.

Given a pipe dream $P$, we write $i\to j$ whenever $i$ has an elbow weakly northwest of $j$. \cref{lem:rectangle} gives us that $i\to j$ implies $i\contactLess{P} j$. 
When $P$ is not acyclic, we can find a cycle ${\bf C}=(i_1, i_2, \ldots, i_m)$ of the contact graph, where $i_1 \to i_2\to\cdots\to i_m\to i_1$. 
We say that $\bf{C}$ is a \defn{reduced cycle} if it is not possible to remove any pipe from $\bf C$ and still have a cycle. 
Given a reduced cycle $\bf C$, we can also assume that $i_1 = \min\{i_1, \dots, i_m\}$.
\vincent{Maybe I would put this cycle story inside the proof.}

To describe the location of crossings and contacts (elbows), we use the coordinate $(c,r)$, where~$r$ is the row and $c$ is the column (ranging from $0$ to~$n$).

@@

\begin{proof}
Let $(c,r)$ denote the position of the crossing~$x$. Since $i$ has an elbow southwest of $x$, we have $r\leq i-1$.
Since $j$ has an elbow northeast of~$x$, there are at most $r-1 \le i-2$ pipes that cross north of $(c,r)$ in column~$c$.
Hence, at least $2$ pipes $<i$ reach the northern border before column~$c$.
Excluding $0$, there is at least a pipe $0<h$ such that $h<i<j$ and $\omega^{-1}(h)< \omega^{-1}(j)<\omega^{-1}(i)$.
This is a $132$ pattern in $\omega$, therefore $\omega$ is not dominant.
\end{proof}

\begin{proof}
For the forward implication, assume that $\omega$ is not dominant.
Let $\greedyPipeDream$ be the greedy pipe dream of $0\omega$. 
\vincent{I believe you wanted the antigreedy here?}
It has no crossing in row 0, and cannot be a partition starting at row 1.
\vincent{I don't like the way it is written here. I would say explicitly the connection between the crossings of the greedy pipe dream and the Rothe diagram of~$\omega$.}
There must be a row $r>1$ with more crossing than in row $r-1$.
\vincent{You mean really inclusionwise on columns?}
Considering the rightmost column $c$ of $P$ with a crossing in row $r$, the pipes $r<k$ are crossing at $(c,r)$, $\omega^{-1}(k)<\omega^{-1}(r)$ and are in contact in position $(c+1,r-1)$.
Let $Q$ be the pipe dream obtained from $\greedyPipeDream$ by the flip of the crossing at $(c,r)$ with the contact at $(c+1,r-1)$.
The contact of $Q$ at $(c,r)$ gives $r\contactLess{Q} k$.
In row $0$ we only have contacts, we have $\omega^{-1}(k)<\omega^{-1}(r)$ and \cref{lem:rectangle} gives that $k \contactLess{Q} r$, hence $Q$ is not acyclic.

For the backward direction, assume we have a pipe dream $P$ that is not acyclic. The contact graph must contain a reduced cycle ${\bf C}=(i, j_1, \ldots, j_m)$ with $i<j_k$ for all $k \in [m]$. 
If $m=1$, then $i$ and $j_1$ must cross , $i\to j_1$ implies there is an elbow of $i$ southwest of the crossing and $j_1\to i$ implies there is an elbow of $j$ northeast of the crossing. \cref{lem:notDominant} implies $\omega$ is not dominant. We consider the case where $m>1$.
For $i\to j_1$, there is an elbow of~$i$ at $(a,b)$ weakly northwest of an elbow of~$j_1$ at $(c_1,d_1)$ an elbow.
Similarly, for $j_m\to i$, there is an elbow of~$j_m$ at $(a_m,b_m)$ weakly northwest of an elbow of~$i$ at~$(c,d)$.
Since the cycle $C$ is reduced, we must have that $(a_m,b_m)$ is not northwest $(c_1,d_1)$. That is $c_1<a_m$ or $d_1<b_m$.

For $d_1<b_m$, we must have $a_m<c_1$ since the pipe $i$ passes northwest of $(c_1,d_1)$ and southeast of $(a_m,d_m)$.
The pipe $j_1$ cannot have an elbow southeast of $(a_m,b_m)$, as otherwise $j_m\to j_1$ would contradict the fact that $\bf C$ is reduced.
The pipes $i<j_1$ must cross in some column $<a_m$ since $j_1$ must remain strictly west of $a_m$ in rows $>b_m$, but the pipe $i$ must have and elbow southeast of $(a_m,b_m)$.
Now, in rows $\le b_m$ the pipe $j_1$ must remain northwest of the pipe $i$. 
It is them impossible to have an elbow in position $(c_1,d_1)$ southeast of $i$ in a row $d_1<b_1$. This is a contradiction and it excludes this case.

\centerline{\includegraphics[scale=.3]{TwoCasesofproof.pdf}}

For $c_1<a_m$, similarly to above,  we must now have $b_m<d_1$ and the pipe $j_m$ cannot have an elbow northwest of $(c_1,d_1)$.
The pipes $i<j_m$ must cross in some column strictly between $c_1$ and $a_m$, since the pipe $j_m$ must remain strictly south of $d_1$ in columns $<c_1$, but the pipe $i$ must have an elbow northwest of $(c_1,d_1)$ and the pipe $j_m$ has an elbow at $(a_m,b_m)$ northwest of $i$.
For the crossing of $i$ and $j_m$,
the pipe $i$ has an elbow at $(a,b)$ southeast of the crossing since $a\le c_1$ and the pipe $j_m$ has an elbow at $(a_m,b_m)$ northeast of the crossing.
\cref{lem:notDominant} implies that $\omega$ is not dominant and conclude the proof.
\end{proof}

@@

\subsection{Linear extensions of facets}
\label{subsec:linearExtensionsFacets}

@@

\begin{remark}
\label{rem:linearExtensionsPartitionSubwordComplexA}
Before proving it, we would like to make the following remarks about theorem.
\begin{enumerate}
\item While the sets~$\linearExtensions(I)$ are order convex, they may not be intervals. See \cref{fig:subwordComplex2,fig:subwordComplex4} (the example of \cref{fig:subwordComplex2} is borrowed from~\cite[Figure~9]{PilaudStump-brickPolytope}).
\item The lower set~$\linearExtensions(Q,\omega)$ may have more than one maximal element, and~$\omega$ is not necessarily maximal in~$\linearExtensions(Q,\omega)$. See \cref{fig:subwordComplex5}.
\item Some of the sets~$\linearExtensions(I)$ may not be included in, nor even meet~$[e,\omega]$. See \cref{fig:subwordComplex5}.
\end{enumerate}
%Note that \cref{fig:subwordComplex4,fig:subwordComplex5} represent type~$A$ subword complexes.
%As in~\cite{PilaudPocchiola, PilaudSantos-brickPolytope, PilaudStump-brickPolytope}, we thus use pseudoline arrangements on sorting networks to represent their facets (and we have represented their contact graphs together).
\end{remark}

>>>>

This paper develops a similar framework for acyclic pipe dreams.
Pipe dreams were introduced by N.~Bergeron and S.~Billey in~\cite{BergeronBilley} to compute Schubert polynomials and later revisited in the context of Gr\"obner geometry by A.~Knutson and E.~Miller~\cite{KnutsonMiller-GroebnerGeometry}, who coined the name \emph{pipe dreams}.
They have important connections and applications to various areas related to Schubert calculus and Schubert varieties~\cite{LascouxSchutzenberger-PolynomesSchubert, LascouxSchutzenberger-SchubertLittlewoudRichardson}. 
A pipe dream is an arrangement of pipes in the triangular shape, each entering along the vertical side and exiting along the horizontal side (see \cref{fig:pipeDreams}).
They are grouped according to their exiting permutation, given by the order in which the pipes appear along the horizontal axis.
The linear extensions of a pipe dream are the permutations of its pipes such that for each contact, the northwest pipe appears before the southeast pipe in the permutation.
The pipe dreams with at least one linear extension are called acyclic and naturally appear in the study of brick polytopes~\cite{PilaudSantos-brickPolytope}.
A flip in a pipe dream exchanges a contact with a crossing between two pipes (see \cref{fig:pipeDreams}), and the flip is increasing when the contact is southwest of the crossing involved in the flip.
A brief recollection on pipe dreams is given in \cref{sec:preliminaries}.

@@

The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j \text{ and } \omega(i)>\omega(j)}$ in matrix notation.
A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrers diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding.
For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~$\{0, 1, \dots, n\}$, and we consider here pipe dreams with pipes indexed by~${0, 1, \dots, n}$.
The $\nu$-Tamari lattice is the increasing flip poset on pipe dreams~$\pipeDreams(0\omega_\nu)$ for some dominant permutation $\omega_\nu$ determined by $\nu$, see~\cite{CeballosPadrolSarmiento}.

The main example is when $\omega \eqdef n \dots 2 1$ is the reverse permutation, see \cref{exm:Tamari1,fig:bijection}.
In this case, the increasing flip poset on~$\pipeDreams(0 \omega)$ is the classical Tamari lattice.
Removing vertex~$0$ from the contact graph of a pipe dream~$ P\in \pipeDreams(0 \omega)$ returns the binary tree corresponding to $P$ (with the leaves removed and the internal nodes labeled in inorder from $1$ to $n$).
The edges of the binary tree are oriented going away from the root of the tree.
In particular, all pipe dreams of~$\pipeDreams(0 \omega)$ are acyclic.
This extends to all dominant permutations as follows.

To prove this result, we will need the following observation.
Note that throughout the proof, the pipes, rows, and column are all indexed by indices ranging from $0$ to~$n$.

@@

\begin{proof}
Let~$r$ be the row and $c$ be the column of the crossing~$x$ (careful that all indices range from~$0$ to~$n$).
Since $i$ has an elbow southwest of $x$, we have $r \le i-1$.
Since $j$ has an elbow northeast of~$x$, there are at most $r-1 \le i-2$ pipes that cross north of $(c,r)$ in column~$c$.
Hence, at least two pipes smaller than~$i$ reach the northern border before column~$c$.
Excluding the pipe~$0$, there is at least one pipe $h > 0$ such that $h < i < j$ and $\omega^{-1}(h) < \omega^{-1}(j) < \omega^{-1}(i)$.
This is a $132$ pattern in the permutation~$\omega$, therefore $\omega$ is not dominant.
\end{proof}

\begin{proof}[Proof of \cref{prob:nuAcyclicProperty}]
For the forward implication, assume that $\omega$ is not dominant, and consider~${i,j,k \in [n]}$ such that~$i < j < k$ while~$\omega^{-1}(i) < \omega^{-1}(k) < \omega^{-1}(j)$, and such that~$k$ is maximal for this property.
Consider any pipe dream~$P \in \pipeDreams(0\omega)$ obtained by capping any pipe dream of~$\pipeDreams(\omega)$ by a top row of contacts.
Because of this top row of contacts, $\omega^{-1}(u) < \omega^{-1}(v)$ implies~$u \contactLess{P} v$.
We now consider the eastmost point~$x$ of pipe~$k$ along row~$i$, and let~$\ell$ denote the other pipe at this point.
We distinguish two cases:
\begin{enumerate}[(i)]
\item Assume first that~$k$ and~$\ell$ cross at~$x$. Since $\omega^{-1}(k) < \omega^{-1}(\ell)$, we have~$k \contactLess{P} \ell$. Moreover, since~$\ell\ne i$, the pipe~$\ell$ must have an elbow west of~$x$ along row~$i$, which implies that~$\ell \contactLess{P} k$. Hence, we have both~$k \contactLess{P} \ell$ and~$\ell \contactLess{P} k$, so that~$P$ is cyclic.
\item Assume now that~$k$ and~$\ell$ touch at~$x$, so that~$k \contactLess{P} \ell$. If~$\ell < k$, then~$k$ and~$\ell$ must cross before~$x$, and flipping~$x$ in~$P$ yields a cyclic facet. If~$\omega^{-1}(\ell) < \omega^{-1}(k)$, then~$\ell \contactLess{P} k$, so that~$P$ is again cyclic. Hence, we can assume that~$k < \ell$ and~$\omega^{-1}(k) < \omega^{-1}(\ell)$. By maximality of~$k$, we obtain that~$j$ and~$\ell$ do not cross, and thus that~$j$ passes northwest of~$x$. We conclude that~$j \contactLess{P} k$. Since~$\omega^{-1}(k) < \omega^{-1}(j)$, we also have~$k \contactLess{P} j$, so that~$P$ is cyclic.
\end{enumerate}

For the backward implication, we introduce a local notation.
We write~$i \to j$ if there is an elbow of pipe~$i$ weakly northwest of an elbow of pipe~$j$.
Note that~$i \to j$ implies~$i \contactLess{P} j$ by \cref{lem:rectangle}.
The reverse direction does not necessarily holds, but clearly holds for cover relations of~$\contactLess{P}$ (as in this case, there is an elbow of~$i$ in contact with an elbow of~$j$).
Assume now that there is a cyclic pipe dream~$P \in \pipeDreams(\omega)$.
We can thus find a cycle~$\b{C}=(i_1, \dots, i_m)$ with~$i_1 \to \dots \to i_m \to i_1$.
We can moreover assume that~$\b{C}$ is reduced in the sense that no subsequence of~$\b{C}$ forms a cycle for~$\to$.

If~$m = 2$, then the pipes~$i_1$ and~$i_2$ must cross at some point~$x$, and $i_1 \to i_2$ implies that $i_1$ has an elbow southeast of~$x$, while $i_2 \to i_1$ implies that $i_2$ has an elbow northwest of~$x$, so that $\omega$ is not dominant by \cref{lem:notDominant}.

Assume now that~$m \ge 3$.
Assume moreover that~$i_1 < i_\ell$ for all~$\ell \in [m]$ (if not, rotate the indices).
Since~$i_m \to i_1$, there is an elbow~$e$ of~$i_m$ weakly northwest of an elbow~$e'$ of~$i_1$.
Since~$i_1 \to i_2$, there is an elbow~$f$ of~$i_1$ weakly northwest of an elbow~$f'$ of~$i_2$.
We now distinguish four cases:
\begin{itemize}
\item If~$e$ is weakly northwest of~$f$, thus of~$f'$, then~$i_m \to i_2$, contradicting the minimality of~$\b{C}$.
\item If~$e$ is northeast of~$f$, since~$i_1 < i_m$, the pipe~$i_m$ starts southwest of~$f$ and ends northeast of~$f$, so that it has an elbow either southeast of~$f$ so that~$i_1 \to i_m$, or northwest of~$f$ thus of~$f'$ so that~$i_m \to i_2$. In both cases, this contradicts the minimality of~$\b{C}$.
\item If~$e$ is southwest of~$f$, as~$i_1$ passes trough~$e'$ and~$f$, and $i_2$ starts south of~$i_1$ (since~$i_1 < i_2$) and passes south of~$f$, we obtain that~$i_2$ must pass south of~$e$, so that $i_m \to i_2$, which contradicts the minimality of~$\b{C}$.
\item If~$e$ is weakly southeast of~$f$, then~$e'$ is southeast of~$f$, which is impossible as $i_1$ has to pass through both~$e'$ and~$f$.
\qedhere
\end{itemize}
\end{proof}

@@

\pagebreak
\subsection{Linear extensions of facets}
\label{subsec:linearExtensionsFacets}

@@

\pagebreak
\begin{remark}
\label{rem:linearExtensionsPartitionSubwordComplexA}
Before proving it, we would like to make the following remarks about theorem.
\begin{enumerate}
\item While the sets~$\linearExtensions(I)$ are order convex, they may not be intervals. See \cref{fig:subwordComplex2,fig:subwordComplex4} (the example of \cref{fig:subwordComplex2} is borrowed from~\cite[Figure~9]{PilaudStump-brickPolytope}).
\item The lower set~$\linearExtensions(Q,\omega)$ may have more than one maximal element, and~$\omega$ is not necessarily maximal in~$\linearExtensions(Q,\omega)$. See \cref{fig:subwordComplex5}.
\item Some of the sets~$\linearExtensions(I)$ may not be included in, nor even meet~$[e,\omega]$. See \cref{fig:subwordComplex5}.
\end{enumerate}
%Note that \cref{fig:subwordComplex4,fig:subwordComplex5} represent type~$A$ subword complexes.
%As in~\cite{PilaudPocchiola, PilaudSantos-brickPolytope, PilaudStump-brickPolytope}, we thus use pseudoline arrangements on sorting networks to represent their facets (and we have represented their contact graphs together).
\end{remark}
